Paper NA

### 2017-03-03 Paper_NA_br.tex
Update 03.03.17

		The average of $\xi$ is higher than $1$ (which means that $V_{SH} > V_{SV}$), but lower than the global average, the latter influenced by the strong $\xi > 1$ oceanic signature.
		% where shearing is more important. 
		Both trends are similar with depth, but the maximum in $\xi$ is slightly deeper than in the global average and the change from $V_{SH} > V_{SV}$ to $V_{SH} < V_{SV}$ is reached at ca. $420km$.

@@

		Our study confirms the presence of a thick lithospheric root in the craton with faster than average $V_S$ down to $250km$ depth, with highest velocities, ca. $4.8km.s^{-1}$, located at depths shallower than  $150km$ (e.g. Figure~\ref{cratoncross}). 
		Laterally, the cratonic root in $V_S$ coincides roughly with precambrian aged crust (see Figure~\ref{3d-VS} from 60 to 150km deep)
		The western edge of the craton follows the RMF (black thick dashed line in Figure \ref{mapgeol}), instead of the western breakup line (See the red thin line in Figure~\ref{mapgeol} which coincides with RMF in Canada, but is diverging westward in the U.S.). While the eastern/south-eastern coincides with the Llano-Grenville Front (See the blue thin line in Figure~\ref{mapgeol}). 
		The northern part ends at the coastline of the Arctic ocean including the arctic archipelago and Greenland.

		These lateral edges are well observed down to $150km$ depth. 
		Below that, the cratonic root retracts toward the center of the archaean continent and is still present beneath Slave, Hearne/Rae and western Superior cratons as shown in Figure \ref{cratoncrossedge}. 
		Everywhere within the craton, a negative gradient of $V_S$ is observed, marking the base of the lithosphere, with a minimum of $\sim 4.6km.s^{-1}$ reached at $\sim 300km$ depth. 
		While the base of the lithosphere (and top of the low velocity zone- LVZ) shows strong topography, the bottom of the LVZ appears largely flat, in particular under the  Slave, Hearne/Rae and western Superior cratons.

		As observed in \cite{babuvska1998age,gung2003global,yuan2014lithospheric}, there are large lateral negative variations $\xi$ down to $\sim 200km$ depth, with regions of reduced radial anisotropy, in contrast to the more uniform, positive $dln(\xi)$ structure in this depth range in the oceans (See figure \ref{3d-Xi}. 
		Below 250km, it is interesting to note that there is a large band of positive anomaly crossing the whole continent (across the US), making the connection between the Pacific and Atlantic mantles. 
		This feature is not present in most of Canada, where negative $dln(\xi)$ structure persist down to 300km.
		Generally, $\xi$ always appear higher than 1. 
		Except between 60 and 100 km, lower than 1 of $\xi$ ($V_{SV} > V_{SH}$) are observed in Mexico, at the southern part of Florida, the southern part of the Superior craton comprising the Mid Continental rifting, the southern part of the Granit Rhyiolite Province comprising the Failed Rifting area, a large area comprising southern Hearne, Medicine hat, northern Wyoming, B.H.T, Cordillera; and finally beneath Baffin Island. This feature can be better observed in Figure \ref{3d-xi-iso-100km} where perturbations are shown according the Isotropic case ($\xi = 1$)

@@

		\paragraph{Wyoming and Medecine Hat}
			The highly reworked Wyoming craton does not show as high anomalies as within the craton nucleus (+2\% intstead of +4-8\%) as shown in figure ~\ref{3d-VS}. 
			It is split into 2 distinctive part. A western part showing low velocities influenced by the active tectonic with a radially positive gradient. 
			In contrast with the eastern part showing higher velocities with a radially negative then positive gradient which marks the western border of the craton nucleus. 
			The depth of negative gradient's minima is variable, from 70 (western part) to 250km (eastern part) deep. 
			The separation between the two areas coincides with the RMF.
			In terms of $\xi$, it is characterized by negative anomalies down to 250km deep. 
			The northern part shows highest anomalies and $V_{SV}$ appears higher than $V_{SH}$ of about 2\% between 60 and 100km. This feature is common with the Medicine Hat province (See Figure \ref{3d-xi-iso-100km}).

		\paragraph{Slave}
			The Slave craton shows typical lithospheric craton signatures with higher than average velocities (See figures \ref{3d-VS} and \ref{cratoncross}), and negative perturbations of $\xi$ (See Figure \ref{3d-VS}. 
			We can note that the average of $\xi$ is the only one higher than the continental mean, compared to the other Archaean provinces (See Figure \ref{1daverage}.
			It shows in the entire region a positive velocity gradient with a maximum of ca. $4.8km.s^{-1}$ associated with a maximum of $\xi$ at ca. $150km$ deep. 
			Deeper, a negative gradient is observed with a minimum of ca. $4.6km.s^{-1}$ at ca. $250km$ deep. 
			The topography of the extrema appears flat.
			Although the Slave lithosphere shows highest velocities at $150km$, above $100km$ velocities are slightly higher than the continental average ($dln(V_S)< 4\%$).

@@

		\paragraph{Western part}
			The western Superior craton shows typical lithospheric craton signatures with highest velocities as shown in figure ~\ref{3d-VS} and in figure~\ref{cratoncross}. In terms of $\xi$, slight negative to zero perturbations are observed at the very western part, while close to the Mid continental rifting, highest negative perturbations indicate that $V_{SV}$ is higher than $V_{SH}$. This last feature is associated the thinning of the cratonic lithosphere when crossing the Grenville front.
			In the entire region there is a positive velocity gradient with a maximum of ca. $4.8km.s^{-1}$  associated with a maximum of $\xi$ at ca. $150km$ deep.

			Deeper a negative gradient is observed with a minimum of ca. $4.6km.s^{-1}$ at ca. $250km$ deep. 
			Opposed to the Slave craton, the topography of the extrema is variable. 
			From SW to NE, the core of maximum velocity can reach beween 70 and 150km of thickness. 
			Beneath the maximum velocity region of ca. $150km$ the negative gradient appears stipper than beneath the $70km$ one, while the topography of the deeper positive gradient appears flat.

			The band of minimum velocity within the cratonic root has similar amplitudes, in terms of velocity, than beneath the younger lithosphere but at shalower depths between $100$ and $150 \: km$. 
			Following the Great Meteor track built by \cite{heaman2000timing}, which crosses the Superior craton, it is interseting to note that the low velocity channel beneath the base of the lithosphere shows lower velocties than usualy observed beneath the other Archaean cratons. 
			This slow anomaly could correspond to an extension the V-shape dent, or divot, observed by \cite{lee1997upper} who explained the feature by the presence of water in the mantle. 
			We observe at $250km$ that the slow anomaly go further inland and crosses the Superior Craton northward until the Hudson bay following the GMT.

		\paragraph{Eastern part}
			Compared to its western part, the structure here shows differences as shown in figure ~\ref{cratoncrossedge} 
			Fast velocities are present down to $150 \: km$ deep but are slower and no perturbations are observed below $200 \: km$. 
			The negative gradient is steeper and the minimum is radially thicker (more than $100 \: km$ thick compare to $\sim 50 \: km$ beneath the craton nucleus). 
			% However at the center, the minimum channel is not continuous and shows higher (still slower than the center of the western Superior) velocities.
			% the gradual trend in velocity is replaced by an almost constant slow anomaly from $150$ to $300km$ deep.

		\paragraph{Trans Hudson Orogen}
			The Trans Hudson Orogen (THO) is a belt of reworked archaean aged region. It refers to a major Paleo Proterozoic tectonic event (between $2.0$ to $1.8Ga$) that welded the Hearne, Wyoming, and Superior archaean blocks to form core of Laurentia \citep{hoffman1988united,whitmeyer2007tectonic}.
			% At shallow depths ($ < 60km$) shows lower velocities than archaean surroundings. 
			It shows typical lithospheric craton signatures with higher than average velocities (See figures \ref{3d-VS} and \ref{cratoncross} associated with negative $\xi$ perturbations. 
			It is interesting to notice that at shallower depths than $100km$, the THO (including the Sask craton) which welded the Superior and Rae/Hearne cratons, shows lighter (i.e, $+4\%$ compared to $+8\%$) positive anomalies than its archaean surrounding. This feature is associated with negative anomalies of $\xi$. While most of the proterozoic root start to thin out at $150km$, the THO at the vicinity of the hudson bay shows as high anomalies as Rae/Hearne and Superior areas associated with zero to positive anomalies in term of $\xi$.

@@

\subsubsection{Surrounding oceanic} This section needs to be redone...

@@

		\caption{\baselineskip 18pt 
		Source and station distribution for the new North American inversion. Black triangles show the seismic stations. Stars are 155 regional events, in addition to 122 teleseismic events. Orange stars indiscate events shallower than $30km$ deep, Yellow stars for deeper than $300km$ and red one are in between. Thick red line indicates the boundaries used for the RegSEM forward simulation, which extends 89◦ by 89◦ horizontally and down to 1600 km and contains all stations and regional events.}

@@

	% dIso for Xi
	\begin{figure}
		\centerline{\includegraphics[width=\linewidth]{figures/Xi_NASEM-iter3-Xi-iso_100km.png}}

		\caption{3D radial anisotropic structure of the continent at 100km compared to the Isotroic case. 
		Map view shows variations with respect to the Isotropic case as $\frac{\xi - 1}{1}$ in \%. 
		Map view is saturated to better show where $V_{SV}$ is stronger than $V_{SH}$.
		Below each map, depth and its regional shear velocity mean is indicated. Minimum and maximum perturbations are also indicated.}
		\label{3d-xi-iso-100km}

>>>>

		The average of $\xi$ is higher than $1$ (which means that $V_{SH} > V_{SV}$), but lower than the global average, the latter influenced by the strong $\xi > 1$ oceanic signature. 
		Both trends are similar with depth, but the maximum in $\xi$ is slightly deeper than in the global average and the change from $V_{SH} > V_{SV}$ to $V_{SH} < V_{SV}$ is reached at $\sim 420 \: km$. 
		At shallower depths than $\sim 200 - 220 \: km$, strongest variations relative the continental average are observed ($+/- 4 \: \text{to} \: 8 \%$, compared to $+/- 1 \: \text{to} \: 3 \%$). 
		Alternations of positive and negative gradient are observed from $50 \: \text{to} \: 400 \: km$ with maxima diminishing along depth. Below than $420 \: km$, negative to isotropic structures with very small lateral variations (i.e $ dln(\xi) < 0.5 \%$) are observed.

		% A negative gradient is regionally observed with its minima at $100 \Km$ of $\sim \xi = 1.02$ evolving to a positive gradient with a maximum of $\xi = 1.03$ located at $150 \: km$. 
		% where shearing is more important.

@@

		Our study confirms the presence of a thick lithospheric root in the craton with faster than average $V_S$ down to $250 \: km$ depth, with highest velocities, ca. $4.7 \: km.s^{-1}$, located at depths shallower than  $150 \: km$ (e.g. Figure~\ref{cratoncross}). 
		Laterally, the cratonic root in $V_S$ coincides roughly with precambrian aged crust (see Figure~\ref{3d-VS} from 60 to 150km deep).
		The western edge of the craton follows the RMF (black thick dashed line in Figure \ref{mapgeol}), instead of the western breakup line (See the red thin line in Figure~\ref{mapgeol} which coincides with RMF in Canada, but is diverging westward in the U.S.). While the eastern/south-eastern coincides with the Llano-Grenville Front (See the blue thin line in Figure~\ref{mapgeol}). 
		The northern part ends at the coastline of the Arctic ocean including the arctic archipelago and Greenland.

		These lateral edges are well observed down to $150 \: km$ depth. 
		Below that, the cratonic root retracts toward the center of the archaean continent and is still present beneath Slave, Hearne/Rae and western Superior cratons as shown in Figure \ref{cratoncrossedge}. 
		Everywhere within the lithospheric core, a negative gradient of $V_S$ is observed, marking the base of the lithosphere \citep{eaton2009elusive}, with a minimum of $\sim 4.6 \: km.s^{-1}$ reached at $\sim 300 \: km$ depth. 
		While the base of the lithosphere (and top of the low velocity zone - LVZ) shows strong topography, the bottom of the LVZ appears largely flat, in particular under the Slave, Hearne/Rae and western Superior cratons.

		As observed in \cite{babuvska1998age,gung2003global,yuan2014lithospheric}, there are large lateral negative variations $\xi$ down to $\sim 200km$ depth, with regions of reduced radial anisotropy, in contrast to the more uniform, positive $dln(\xi)$ structure in this depth range in the oceans (See figure \ref{3d-Xi}).

		{\color{red} I would like to point that we dont see the same features as in \cite{marone2007three} with the craton showing same $\xi \: (> 1)$ than oceanic lithopshhere, while western active part and greenvile-appalachian area shows reduced $\xi < 1$. Different causes to me: starting model, I need to compare my and her starting model, or simply the lateral parameterisation (8deg (federica) compared to 2deg (Me)).}

		Below 250km, it is interesting to note that there is a large band of positive $dln(\xi)$ crossing the whole continent (across the US), making the connection between the Pacific and Atlantic mantles. 
		This feature is not present in most of Canada, where negative $dln(\xi)$ structure persist down to 300km.
		Generally, $\xi$ always appears higher than 1. 
		Except between 60 and 100 km, lower than 1 of $\xi$ ($V_{SV} > V_{SH}$) are observed in Mexico, at the southern part of Florida, the southern part of the Superior craton comprising the Mid Continental rifting, the southern part of the Granit Rhyiolite Province comprising the Failed Rifting area, a large area comprising southern Hearne, Medicine hat, northern Wyoming, B.H.T, Cordillera; and finally beneath Baffin Island. This feature can be better observed in Figure \ref{3d-xi-iso-100km} where perturbations are shown according the Isotropic case ($\xi = 1$)

@@

		\paragraph{Wyoming and Medecine Hat}
			The Wyoming craton was strongly affected by a regional metamorphic event related to the Trans-Hudson orogen, which juxtaposed a number of Archean cratons during the Paleoproterozoic development of Laurentia \citep{hoffman1989precambrian}.
			This highly reworked craton does not show as high anomalies as within the craton nucleus (+2\% instead of +4 \:\text{to}\: 8\%) as shown in figure ~\ref{3d-VS}. 
			It is split into 2 distinctive part. A western part showing low velocities influenced by the active tectonic with a radially positive gradient in terms of $V_S$. 
			In contrast with the eastern part showing higher velocities with a radially negative then positive gradient which marks the western border of the craton nucleus. 
			The depth of negative gradient's minima is variable, from 70 (western part) to 250km (eastern part) deep. 
			The separation between the two areas coincides with the RMF.
			In terms of $\xi$, it is characterized by negative anomalies down to 250km deep. 
			The northern part shows highest negative $dln(\xi)$ and $V_{SV}$ appears higher than $V_{SH}$ of about 2\% between 60 and 100km. This feature is common with the Medicine Hat province (See Figure \ref{3d-xi-iso-100km}).

		\paragraph{Slave}
			The Slave craton shows typical lithospheric craton signatures with higher than average velocities (See figures \ref{3d-VS} and \ref{cratoncross}), and negative perturbations of $\xi$ (See Figure \ref{3d-Xi}). 
			We can note that the average of $\xi$ is the only one higher than the continental mean, compared to the other Archaean provinces (See Figure \ref{1daverage}).
			It shows in the entire region a positive velocity gradient with a maximum of ca. $4.7 \: km.s^{-1}$ associated with a maximum of $\xi$ at ca. $150 \: km$ deep. However we don't observe, both the Central Slave Mantle Conductor and the $4.8 \: km.s^{-1} \: V_S$ profile extending from $160 \: \text{to} \: 250 \: km$, obtained from joint inversion of receiver function, surface wave dispersion curve, and magnetotelluric data of \cite{moorkamp2010joint}. 
			Deeper, a negative gradient is observed with a minimum of ca. $4.65 \: km.s^{-1}$ at ca. $250 - 270\: km$ deep in agreement with \cite{chen2007new}. Not sure I should cite error bars are .0.2kms-1...
			Although the Slave lithosphere shows highest velocities at $150 \: km$, above $100 \: km$ velocities are slightly higher than the continental average ($dln(V_S)< 4\%$).

			Je voudrais ici mettre en lien avec les mesures de kimberlites/ eclogites de \cite{faure2011seismic,heaman2003timing} Aulbach et al., 2007, Jacob 2004 pour montrer que leur mesures estimant la profondeur de la pLAB (200 a 220) peut ne pas etre compatible avec les mesures tomo car leurs mesures datent d'entre 650Ma et 50Ma. Il peut y avoir eu un epaississement du mantle keel, mais je n'ai pas encore trouve d'arguments favoarbles... Ou alors que ce genre d'information peut etre mieux contraint via l'aniso azim mais tous leurs papiers comparent la LAB sismique avec Vs only.

			% Two firm LAB estimates come from MT. The star data point in the centre of the Slave Craton in northern Canada (Fig. 2e) is the LAB estimate of Moorkamp et al. (2010), derived from joint inver- sion of receiver function data, surface wave dispersion curve data, and magnetotelluric data from the EKTN POLARIS station. This eLAB + sLABrf + sLABsw value of 230 km (220–235 km) exceeds by 20–30 km the greatest pLAB values on the Slave Craton of 202 km for the Neoproterozoic Anuri pipe in NW Slave, 208 km from the Jurassic-aged Gahcho Kue pipe (formerly known as Kennedy Lake) in SW Slave, and the 202 km pLAB from the Eocene-aged Diavik pipe in the central Slave (Snyder and Grütter 2010). It is though well within the error limits of the poorly-resolved sLABsw esti- mate of 220 ± 60 km of Chen et al. (2007) based on only Rayleigh wave inversion from the same station (EKTN); here, the joint in- version of Moorkamp et al. (2010) has served to reduce the uncer- tainty considerably. 
			{\color{red}Either the pLAB estimates are too low by 20–30 km, which is likely, or the lithosphere has grown by that order over the last 50 million years since the Eocene kimberlite eruptions in the central Slave, which is less likely, or aspects of both are operating.} \cite{jones2014electrical}

@@

		\paragraph{Western Ontario part}
			The western Superior craton shows typical lithospheric craton signatures with highest velocities as shown in figure ~\ref{3d-VS} and in figure~\ref{cratoncross}. In terms of $\xi$, slight negative to zero perturbations are observed at the very western part, while close to the Mid continental rifting, highest negative perturbations indicate that $V_{SV}$ is higher than $V_{SH}$. This last feature is associated with the thinning of the cratonic lithosphere when crossing the Grenville front as observed by \cite{darbyshire2007new}.
			In the entire region there is a positive velocity gradient with a maximum of ca. $4.7 \: km.s^{-1}$  associated with a maximum of $\xi$ at ca. $150km$ deep.

			Deeper a negative gradient is observed with a minimum of ca. $4.6km.s^{-1}$ at ca. $250km$ deep. 
			Opposed to the Slave craton, the topography of these extrema is variable. 
			From SW to NE, the core of maximum velocity can reach beween 70 and 150km of thickness. 
			Beneath the maximum velocity region of ca. $150km$ the negative gradient appears stipper than beneath the $70km$ one, while the topography of the deeper positive gradient appears flat.

			The band of minimum velocity within the cratonic root has similar amplitudes, in terms of velocity, than beneath the younger lithosphere but at shalower depths between $100$ and $150 \: km$. 
			Following the Great Meteor track built by \cite{heaman2000timing}, which crosses the Superior craton, it is interseting to note that the low velocity channel beneath the base of the lithosphere shows lower velocties than usualy observed beneath the other Archaean cratons. 
			Xenoliths analysis from kimberlites from Kirkaland Lake (south eastern part of the western Superior) exhibit some evidence of diamonds, i.e, the base of the lithosphere was deeper than 170 km at the time of kimberlitic eruption \citep{jones2014electrical} estimated at $160 - 152 \: Ma$ by \cite{heaman2000timing,heaman2003timing}. 
			This can be interpreted as either the lithosphere is today thick, and the lower lithosphere was not sampled by xenolith entrainment in the kimberlitic magmas (but nevertheless diamonds were entrained), or the lowermost lithosphere has been eroded away since eruption, possibly by the movement of the North American plate over the Great Meteor hotspot \citep{heaman2000timing,jones2014electrical}. The latter being in agreement with our of observations.

			% This slow anomaly could correspond to an extension the V-shape dent, or divot, observed by \cite{lee1997upper} who explained the feature by the presence of water in the mantle. 
			% We observe at $250km$ that the slow anomaly go further inland and crosses the Superior Craton northward until the Hudson bay following the GMT.

		\paragraph{Eastern Quebec part}
			Compared to its western part, the structure here shows differences as shown in figure ~\ref{cratoncrossedge} 
			Fast velocities are present down to $150 \: km$ deep but are slower and no perturbations are observed below $200 \: km$. 
			The negative gradient is steeper and the minimum is radially thicker (more than $100 \: km$ thick compare to $\sim 50 \: km$ beneath the craton nucleus). 
			% However at the center, the minimum channel is not continuous and shows higher (still slower than the center of the western Superior) velocities.
			% the gradual trend in velocity is replaced by an almost constant slow anomaly from $150$ to $300km$ deep.

		\paragraph{Trans Hudson Orogen}
			The Trans Hudson Orogen (THO) is a belt of reworked archaean aged region. It refers to a major Paleo Proterozoic tectonic event (between $2.0$ to $1.8Ga$) that welded the Hearne, Wyoming, and Superior archaean blocks to form core of Laurentia \citep{hoffman1988united,whitmeyer2007tectonic}.
			% At shallow depths ($ < 60km$) shows lower velocities than archaean surroundings. 
			It shows typical lithospheric craton signatures with higher than average velocities (See figures \ref{3d-VS} and \ref{cratoncross} associated with negative $\xi$ perturbations. 
			It is interesting to notice that at shallower depths than $100km$, the THO which welded the Superior and Rae/Hearne cratons, shows lighter (i.e, $+4\%$ compared to $+8\%$) positive anomalies than its archaean surrounding. This feature is associated with negative anomalies of $\xi$. While most of the proterozoic root start to thin out at $150km$, the THO at the vicinity of the hudson bay shows as high anomalies as Rae/Hearne and Superior areas associated with zero to positive anomalies in term of $\xi$. 
			The THO comprises the archaean aged Sask craton (See Figure \ref{mapgeol}. 
			The mantle beneath this exotic craton is interpreted to have been detached and replaced by both the Superior and the Hearne mantle lithosphere during final collision of the orogen \citep{nemeth2005lithospheric}. 
			This interpretation, according to \cite{faure2011seismic} can be extrapolated farther to the east and may explain why Trans-Hudson orogen has no imprint in the upper mantle.

@@

\subsubsection{Surrounding oceanic}

@@

		\caption{\baselineskip 18pt 
		Source and station distribution for the new North American inversion. Black triangles show the seismic stations. Stars are 155 regional events, in addition to 122 teleseismic events. Orange stars indiscate events shallower than 30~km deep, yellow stars for deeper than 300~km and red one are in between. Thick red line indicates the boundaries used for the RegSEM forward simulation, which extends 89◦ by 89◦ horizontally and down to 1600~km and contains all stations and regional events.}

@@

	% dIso for Xi
	\begin{figure}
		\begin{minipage}{0.5\linewidth}
			\centerline{\includegraphics[width=\linewidth]{figures/Xi_NASEM-iter3-Xi-iso_60km.png}}
		\end{minipage}
		\hfill
		\begin{minipage}{0.5\linewidth}
			\centerline{\includegraphics[width=\linewidth]{figures/Xi_NASEM-iter3-Xi-iso_100km.png}}
		\end{minipage}
		% \begin{minipage}{0.5\linewidth}
		% 	\centerline{\includegraphics[width=\linewidth]{figures/Xi_NASEM-iter3-Xi-iso_150km.png}}
		% \end{minipage}

		\caption{\baselineskip 18pt 
		3D radial anisotropic structure of the continent at 60 and 100 \: km compared to the Isotropic case. 
		Map view shows variations with respect to the Isotropic case as $\frac{\xi - 1}{1}$ in \%. 
		Map view is saturated to better show where $V_{SV}$ is stronger than $V_{SH}$.
		Below each map, depth and its regional shear velocity mean is indicated. Minimum and maximum perturbations are also indicated.}
		\label{3d-xi-iso-100km}
